A Quantitative Genetic Linked Selection Model of the Negative Selection Process in Humans

### 2023-07-12 manuscript/manuscript.tex
ms changes

Similar to previous work, we aim to characterize the negative selection process
through the signal left on nucleotide diversity at linked sites. Although
nucleotide diversity is a less-informative statistic than the full site
frequency spectrum, it is much easier to derive theoretic models of how
selection alters expected pairwise coalescence times at linked sites than to
describe the entire genealogy. Still, inferences drawn from the linked
selection signal about selective processes are only as accurate as the
underlying theoretic model. Our work is motivated by a well-known limitation of
classic background selection theory: it only accurately predicts levels of
linked diversity when new mutations are strongly deleterious
\parencite{Charlesworth1993-gb,McVean2000-bt,Good2013-lp,Gordo2002-dr}.
However, developing linked selection theory that accommodates weak selection
has been challenging problem \parencite{Good2014-yz,Haigh1978-gt,Higgs1995-xc},
and it is worthwhile to develop in intuition for how the classic BGS model
works, and how quantitative genetic models of linked selection can alleviate
these problems and be extended to genome-wide inference.

Both the classic background selection and SC16 models imagine a negative
selection process where a continual influx of deleterious mutations flow into
the population at a rate of $\mu$ per basepair per generation in a conserved
region of $L$ basepairs, such that the region-wide per generation mutation rate
is $U = \mu L$. Each mutation imposes an additive selective cost of $s$ in
heterozygotes and $2s$ in homozygotes. The classic background selection model
only considers mutations that are sufficiently deleterious that they are
destined to extinction
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1994-oh}. In
this case, the genealogy is well-approximated by a neutral coalescence process
with a rescaled effective population size of $N_0 = N \exp(-\nicefrac{U}{s})$.
This approximation works reasonably well for two reasons (though see
\cite{Cvijovic2018-vd,Walczak2012-fi,Nicolaisen2012-vs}). First, when selection
is strong, the number of deleterious mutations per haplotype has a stationary
Poisson distribution with an average equal to the equilibrium under the
deterministic mutation-selection model \parencite{Haldane1927-ga}. Since
lineages carrying strongly deleterious mutations cannot contribute to the
genealogy in the long run, coalescence events only occur in the fraction
$\exp(-\nicefrac{U}{s})$ of the population free of mutations. Second, the time
it takes for a present-day lineage to trace their ancestry back to a
mutation-free ancestor (the ``delay phase") is negligible compared to the
coalescence timescale of $N_0$ among these ancestors (the ``coalescence
phase"), and can be ignored (\cite{Durrett2008-ql}, p. 213;
\cite{Good2014-yz}).

By contrast, the frequency trajectories of weakly deleterious mutations are
strongly impacted by stochastic perturbations, so deterministic models no
longer provide a good approximation. When genetic drift is the only source of
these randomness in allele frequency change, the relevant scale of stochastic
perturbations is determined by the drift-effective population size $N_e$
\parencite{Ohta1971-gq,Ohta1992-yi}. Under weak selection ($2N_e s \le 1$), the
stochastic perturbations are on a scale equal to or greater than the changes
due to selection. Then, weakly deleterious alleles can by chance drift up to
intermediate frequencies before their eventual loss or fixation. Classic BGS
breaks down in this domain, since the ``delay phase" is no longer negligible
and genealogies are no longer well-approximated by a neutral coalescence
process \parencite{Przeworski1999-mb,OFallon2010-my,Higgs1995-xc}. Moreover,
the distribution of the number of deleterious mutations (and its corresponding
fitness distribution) are no longer stationary and become a traveling wave
\parencite{Rouzine2008-qz,Good2013-lp,Gessler1995-hz} towards reduced
population fitness (since beneficial and back mutations are ignored). When this
occurs, individuals in the least-loaded class are stochastically lost, fixing
those mutations with a click of ``Muller's ratchet" \parencite{Muller1964-ki}.
Determining both the rate of Muller's ratchet
\parencite{Haigh1978-gt,Gordo2002-dr,Gessler1995-hz} and the shape of
genealogies under weak selection have been stubborn open problems in
evolutionary genetics.

To further complicate matters, when selection is weak, the stochastic
perturbations from selection at neighboring sites generates another source of
randomness in allele frequency change known as genetic \emph{draft}
\parencite{Neher2013-dz}. An allele experiences genetic draft when it becomes
randomly associated with a fitness background, which stochastically perturbs
its trajectory across generations. Like genetic drift, perturbations due to
draft are directionless but overall increase the variance in allele frequency
change. Unlike drift, stochastic perturbations caused by draft are
autocorrelated across generations for as long an allele does not recombine or
segregate off its fitness background and the selective environment does not
change \parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. When
selection is strong, draft can generate multiple-merger coalescences that
create discontinuities in allele frequency trajectories, making diffusion
approximations ill-suited \parencite{Gillespie2000-mh,Der2011-it,Neher2013-dz}.
However, when selection is weak, draft is well-approximated as an increase in
the rate of drift (or equivalently, a reduction in effective population size).

Quantitative genetics models of linked selection formulate the impact of weak
draft in terms of additive genetic fitness variance, which can more readily
accommodate weak selection. This idea derives from a key insight from Robertson
(\citeyear{Robertson1961-ho}): in the long-run, the effect of weak draft on
neutral allele frequencies acts like an increase in the variance in offspring
number. However, unlike non-heritable variation in offspring number, the
heritable variation that generates draft is autocorrelated across generations.
At the individual level Robertson considered, this is because offspring from
large families tend to beget many descendents themselves (and likewise with
small families). The same autocorrelation occurs at the genomic level, as the
perturbations to a neutral allele's trajectory from its particular fitness
background tend to occur in the same direction across generations as long as
the linkage exists and the selective environment is constant. Thus, the total
additional heritable variation in offspring number is the product of the
equilibrium additive fitness variation $V_A$ in a generation and an inflation
factor $Q_t^2$ due to the build up of autocorrelation to generation $t$. We let
$Q^2 := Q_t^2$ as $t \to \infty$. Intuitively, the product $V_A Q^2$ represents
the expected total variance in reproductive success a neutral mutation
experiences over its lifetime in a system with weak selection draft.

@@

>>>>

Similar to previous work, we aim to characterize the negative selection process
through the signal left on nucleotide diversity at linked sites. Although
nucleotide diversity is a less-informative statistic than the full site
frequency spectrum, it is much easier to derive theoretic models of how linked
selection alters expected pairwise coalescence times than to describe the
entire genealogy. Still, inferences drawn from the linked selection signal
about selective processes are only as accurate as the underlying theoretic
model. Our work is motivated by a well-known limitation of classic background
selection theory: it only accurately predicts levels of linked diversity when
new mutations are strongly deleterious
\parencite{Charlesworth1993-gb,McVean2000-bt,Good2013-lp,Gordo2002-dr}.
Unfortunately, modifying background selection theory so that it accommodates
weak selection has been challenging problem
\parencite{Good2014-yz,Haigh1978-gt,Higgs1995-xc}. Here, we develop intuition
for why the classic BGS model works well for strong selection but not weak, and
how formulating selection in terms of the additive fitness variance it creates
can alleviate these problems.

Throughout, we imagine a negative selection process where a continual influx of
deleterious mutations occur at a rate of $\mu$ per basepair per generation in a
conserved region of $L$ basepairs, such that the region-wide per generation
mutation rate is $U = \mu L$. Each mutation imposes a selective cost of $s$ in
heterozygotes and $2s$ in homozygotes, and fitness effects are multiplicative
across sites. When selection against a new mutation is strong, the stochastic
perturbations it experiences due to random genetic drift are weak relative to
the frequency changes caused by selection. In this case, every deleterious
mutation is destined to eventual extinction, and only the fraction of lineages
without any mutations contribute to the long-run ancestry of future
populations. The classic background selection approximates this process by
rescaling the neutral coalescent with an effective population size of $N_0 = N
\exp(-\nicefrac{U}{s})$
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1994-oh}.
This approximation works well for two reasons (though see
\cite{Cvijovic2018-vd,Walczak2012-fi,Nicolaisen2012-vs}). First, he number of
mutations per haplotype has has a stationary Poisson distribution with an
average equal to the deterministic mutation-selection equilibrium
\parencite{Haldane1927-ga}. Second, the time it takes for a present-day lineage
to trace their ancestry back to a mutation-free ancestor (the ``delay phase")
is negligible compared to the coalescence timescale of $N_0$ among
mutation-free ancestors (\cite{Durrett2008-ql}, p. 213; \cite{Good2014-yz}).

By contrast, when selection is weak, the frequency trajectories of both neutral
and selected sites are strongly impacted by stochastic perturbations from two
sources. First, genetic drift is due to Mendelian segregation and non-heritable
fitness differences; the magnitude of its perturbations are set by the
\emph{drift-effective} population size, $N_e$. Second, genetic draft occurs
when an allele becomes randomly associated with a fitness background that
perturbs its trajectory across generations \parencite{Neher2013-dz}. Like
genetic drift, the perturbations due to draft are directionless but overall
increase the variance in allele frequency change. Unlike drift, the impact of
the stochastic perturbations due to draft are autocorrelated and build up over
generations until the fitness background recombines off
\parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. While draft can
generate multiple-merger coalescences that create discontinuities in allele
frequency trajectories, making diffusion approximations ill-suited
\parencite{Gillespie2000-mh,Der2011-it,Neher2013-dz}. However, when selection
is weak, draft is well-approximated as an increase in the rate of drift (or
equivalently, a reduction in effective population size). The stochastic
perturbations between selecting sites creates Hill--Robertson interference
\parencite{Hill1966-kd}.

The stochastic perturbations from genetic drift and draft impact both
genealogies and selection dynamics in complex ways, which has made finding
analytic theory predicting their joint effects difficult. When drift acts
alone, the relevant scale of perturbations is determined by the
\emph{drift-effective population size} $N_e$
\parencite{Ohta1971-gq,Ohta1992-yi}. When the random perturbations due to drift
are stronger than the changes due to selection ($2N_es \le 1$), weakly
deleterious mutations can drift up to intermediate frequencies before their
eventual loss or fixation. The assumptions of classic BGS suddenly breakdown in
two ways. First, the ``delay phase" is no longer negligible and genealogies are
no longer well-approximated by a rescaled neutral coalescence process
\parencite{Przeworski1999-mb,OFallon2010-my,Higgs1995-xc}. Second, the
distribution of the number of deleterious mutations (and its corresponding
fitness distribution) is no longer a stationary Poisson distribution, instead
becoming a traveling wave \parencite{Rouzine2008-qz,Good2013-lp,Gessler1995-hz}
towards reduced population fitness (since we ignore beneficial mutations). When
this occurs, individuals in the least-loaded class are stochastically lost,
fixing those mutations with a click of ``Muller's ratchet"
\parencite{Muller1964-ki,Charlesworth1997-qn}. Unfortunately, determining the
rate of Muller's ratchet is another difficult, open problem
\parencite{Haigh1978-gt,Gordo2002-dr,Gessler1995-hz} related to Hill--Robertson
interference \parencite{Felsenstein1974-xm}.

A recent model from \textcite{Santiago2016-mu} addresses these limitations by
formulating the negative selection process in terms of the equilibrium fitness
variance $V_A$ under mutation, drift, and draft.

Since their earlier work has
related the additive fitness variance in a region to the draft-effective
population experienced by a neutral allele arbitrarily far away
\parencite{Santiago1998-bs}, this can be

HERE

that derives the effective population
size experienced by a neutral allele when there is heritable fitness in the
population. This theory is in essence a model for the magnitude of stochastic
perturbations (what we think of as a \emph{draft-effective} population size)
when heritable fitness variation generates weak genetic draft. As mentioned,
perturbations due to draft are autocorrelated through time; at the individual
level Robertson considered, this is because offspring from large families tend
to beget many descendents themselves (and likewise with small families).
However, the same autocorrelation occurs at the genomic level with linkage
\parencite{Santiago1998-bs}, as the perturbations to a neutral allele's
trajectory from its particular fitness background tend to occur in the same
direction across generations until recombination.

The benefit of this quantitative genetics formulation of linked selection is
two-fold. First, as we will see, Santiago and Caballero's
(\citeyear{Santiago2016-mu}) model for the additive fitness variance under both
strong and weak negative selection is quite accurate. Second, these models of
linked selection can 
They do this through

Thus, the total additional
heritable variation in offspring number is the product of the equilibrium
additive fitness variation $V_A$ in a generation and an inflation factor
$Q_t^2$ due to the build up of autocorrelation to generation $t$. We let $Q^2
:= Q_t^2$ as $t \to \infty$. Intuitively, the product $V_A Q^2$ represents the
expected total variance in reproductive success a neutral mutation experiences
over its lifetime in a system with weak selection draft.

@@

draft limits discussion - we were heavily constrained by simulations / regions
etc
